Real-Time Variational Density Forecasts

### 2017-03-05 Confirmation/presentations.Rnw
presentation tweaks

\begin{frame}
\section{Motivation}
\frametitle{Motivation}
\begin{itemize}
\item There is an increasing demand for forecasts to be of the entire predictive density of the variable of interest.
\item Many time series of interest can operate on a very short time-frame.
\item We aim to provide density forecasts for these high frequency time-series by updating posterior and predictive density estimates as each data point is observed.
\item The algorithm use should scale to multivariate observations driven by complex models with large datasets.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Bayesian Computation Techniques}
A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
\begin{itemize}
\item Converges in distribution to the posterior density with enough iterations.
\item Can be very slow, and is hard to parallelise.
\item Scales badly to large datasets.
\end{itemize}
Interest is developing in Bayesian approximations, such as Variational Bayes (VB). VB:
\begin{itemize}
\item Is typically orders of magnitude faster than MCMC.
\item Scales very well to large datasets with subsampling and parallelisation.
\item Has no theoretical guarantee to converge to the posterior.
\end{itemize}
\end{frame}

\begin{frame}
\section{Markov Chain Monte Carlo}
\frametitle{Markov Chain Monte Carlo}
\begin{itemize}
\item A Gibbs MCMC sampler iteratively samples the full conditional distributions:
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
\item Requires convergence, which can be extremely slow.
\item A predictive density for the next data point may not be available before it is observed.
\end{itemize}
\end{frame}

>>>>

\begin{frame}
\section{Motivation}
\frametitle{Motivation}
\begin{itemize}
\item There is an increasing demand for forecasts of the entire predictive density of the variable of interest.
\item Many time series of interest can operate on a very short time-frame.
\item We aim to provide density forecasts for these high frequency time-series by updating posterior and predictive density estimates as each data point is observed.
\item The algorithm used should scale to multivariate observations driven by complex models with large datasets.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Bayesian Computation Techniques}
A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
\begin{itemize}
\item Converges to the posterior density.
\item Can be very slow.
\item Scales badly to large datasets.
\end{itemize}
Interest is developing in Bayesian approximations, such as Variational Bayes (VB). VB:
\begin{itemize}
\item Significantly faster than MCMC.
\item Scales to large datasets.
\item No theoretical guarantee to converge to the posterior.
\end{itemize}
\end{frame}

\begin{frame}
\section{Markov Chain Monte Carlo}
\frametitle{Markov Chain Monte Carlo}
\begin{itemize}
\item A Gibbs MCMC sampler iteratively samples the full conditional distributions:
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
\item Requires many samples to converge, which can be extremely slow.
\item A predictive density for the next data point may not be available before it is observed.
\end{itemize}
\end{frame}

### 2017-03-05 Confirmation/presentations.tex
presentation tweaks

@@

@@

\begin{frame}
\frametitle{Bayesian Computation Techniques}
A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
\begin{itemize}
\item Converges in distribution to the posterior density with enough iterations.
\item Can be very slow, and is hard to parallelise.
\item Scales badly to large datasets.
\end{itemize}
Interest is developing in Bayesian approximations, such as Variational Bayes (VB). VB:
\begin{itemize}
\item Is typically orders of magnitude faster than MCMC.
\item Scales very well to large datasets with subsampling and parallelisation.
\item Has no theoretical guarantee to converge to the posterior.
\end{itemize}
\end{frame}

\begin{frame}
\section{Markov Chain Monte Carlo}
\frametitle{Markov Chain Monte Carlo}
\begin{itemize}
\item A Gibbs MCMC sampler iteratively samples the full conditional distributions:
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
\item Requires convergence, which can be extremely slow.
\item A predictive density for the next data point may not be available before it is observed.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Deriving Gibbs Full Conditional Distributions}
\begin{itemize}
\item Let $y_t \overset{iid}{\sim} \mathcal{N}(\mu, \sigma^2) \mbox{ for } t = 1, 2, \dots, T$. 
\item Use priors $p(\sigma^2) \sim IG(\mbox{shape} = \alpha, \mbox{scale} = \beta), p(\mu | \sigma) \sim \mathcal{N}(\gamma, \sigma^2/\tau)$.
\item Gibbs MCMC requires the distribution $p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T})$. 
\item Step One: Write the full log joint distribution,
\end{itemize}
\begin{align}
\log(p(y_{1:T}, \mu, \sigma^2)) &=  -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} \nonumber \\
&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2}  + c. \nonumber 
\end{align}
\end{frame}

\begin{frame}
\frametitle{Deriving Gibbs Full Conditional Distributions}
\begin{itemize}
\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, y_{1:T}))$,
\begin{equation}
\log(p(\mu | y_{1:T}, \sigma^2)) =  - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c. \nonumber
\end{equation}
\item Step Three: Condition on the previous draw of $\sigma^2$ to form $\log(p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T}))$,
\begin{equation}
\log(p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T})) = - \frac{\sum_{t}(y_t - \mu_{(i)})^2}{2\sigma_{(i)}^2} - \frac{\tau(\mu_{(i)} - \gamma)^2}{2\sigma_{(i)}^2}  + c. \nonumber
\end{equation}
\item Step Four: Manipulate equation until a distribution kernel is recognised. If unrecognised, add a Metropolis-Hastings step.\
\item Step Five: Repeat for $p(\sigma^2_{(i)} | \mu_{(i-1)}, y_{1:T}).$
\end{itemize}
\end{frame}

\begin{frame}
\section{Variational Bayes}
\frametitle{Variational Bayes}
\begin{itemize}
\item Key Idea: Replace the intractable $p(\boldsymbol{\theta} | y_{1:T})$ with a tractable approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
\item How: Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions that minimises a well defined divergence between $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ and $p(\boldsymbol{\theta} | y_{1:T})$. 
\item The classic choice is the Kullback-Leibler divergence from $q$ to $p$,
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
\end{equation}
\item (\ref{KL-def}) is non-negative and equals zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Evidence Lower Bound}
\begin{itemize}
\item Rewrite (\ref{KL-def}) as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} |y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}),
\end{equation}
where
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \mathbb{E}_{q} \left( \ln (p(y_{1:T},\boldsymbol{\theta})) - \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) \right).
\end{equation}
\item $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is refered to as the Evidence Lower Bound (ELBO).
\item Maximising the ELBO with respect to $q \iff$ Minimising the KL Divergence with respect to $q$.
\item There are two tasks: 
  \begin{enumerate}
  \item Choose a functional form for $q$. 
  \item Optimise auxillary parameter vector $\boldsymbol{\lambda}$. 
  \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Mean Field Variational Bayes}
\begin{itemize}
\item Early Variational Bayes implementation pioneered by \citet{Jordan1999} and \citet{Ghahramani2000}.
\item Restricts $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the factorisable family,
\begin{equation}
\label{mf1}
q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
\end{equation}
\item Restricts the model to the exponential family with conjugate priors.
\item Provides the optimal distribution for each $q_i(\theta_i | \boldsymbol{\lambda_i})$.
\item Maximise the ELBO by setting each factor to
\begin{equation}
\label{mf2}
q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{\prod_{j \neq i} q_i(\theta_i | \boldsymbol{\lambda_i})} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
\end{equation}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Deriving $q$ distributions}
\begin{itemize}
\item (\ref{mf2}) is applied to the Normal-Inverse Gamma model to find $q(\mu | \boldsymbol{\lambda}_1)$.
\item Step One: Write the full log joint distribution,
\begin{align}
\log(p(y_{1:T}, \mu, \sigma^2)) &= -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} \nonumber \\
&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2} + c. \nonumber 
\end{align}
\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, y_{1:T})),$
\begin{equation}
\log(p(\mu | y_{1:T}, \sigma^2)) = - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c. \nonumber
\end{equation}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Deriving $q$ distributions}
\begin{itemize}
\item Step Three: Take the expectation with respect to $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to form $\log(q(\mu | \boldsymbol{\lambda}_1)),$
\begin{equation}
\log(q(\mu | \boldsymbol{\lambda}_1)) = - 1/2 \mathbb{E}_{q(\sigma^2)} [\sigma^{-2}] \left( \sum_{t}(y_t - \mu)^2 + \tau(\mu - \gamma)^2 \right)  + c. \nonumber
\end{equation}
\item Step Four: Manipulate equation until a distribution kernel is recognised. If unrecognised, add a secondary approximation.
\item Step Five: Repeat for $q(\sigma^2 | \boldsymbol{\lambda}_2)$.
\item In this case $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim IG(\tilde{\alpha}, \tilde{\beta})$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Deriving $q$ distributions}
\begin{itemize}
\item The auxillary parameter vector is $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$, which maximises the ELBO when
\begin{align}
\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}[\sigma^{-2}]\right )^{-1} \label{mf6} \\
\tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
\tilde{\beta} &= \beta + 1/2\bigg((\tau + T)\mathbb{E}_{q(\mu)}[\mu^2 ]  \nonumber \\ 
&- 2 \mathbb{E}_{q(\mu)}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \bigg). \label{mf8}
\end{align}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Deriving $q$ distributions}
\begin{itemize}
\item Substitute in expectations to (\ref{mf6}) and (\ref{mf8}):
\begin{align}
\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
\end{align}
\item The circular dependence in (\ref{mf9}) and (\ref{mf10}) and can be resolved with a coordinate ascent algorithm.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Fit to simulated data}
\begin{itemize}
\item Marginal posterior densities for the Normal Inverse Gamma model with $N = 100, \mu = 2$ and $\sigma^2 = 1$; hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. 
\end{itemize}
\begin{figure}[h]
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{norminvg}
\caption{The analytical true density (black), the Monte Carlo (red) and the MFVB approximation (blue). MFVB converged in same time as required to take three MCMC samples of each parameter.}
\label{fig:norminvg}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Stochastic Variational Bayes}
\begin{itemize}
\item \citet{Paisley2012} and \citet{Ranganath2014} adapted a gradient ascent algorithm to maximise the ELBO.
\item Gradient ascent returns the optimal $\boldsymbol{\lambda}$ for a specified functional form $q$.
\item Only requires Monte-Carlo estimates of $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$,
\begin{align}
\label{SVB}
\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) &\approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda})] \nonumber \\
&\times \big(\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}))\big) 
\end{align}
where $\boldsymbol{\theta}_s$ is simulated from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ for $s = 1, 2, \dots, S$. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Gradient Ascent}
\begin{itemize}
\item Make steps of the form
\begin{equation}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} f(\lambda^{(m)})
\end{equation}
until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$, with $\rho^{(m)}, m = 1, 2, \dots$ that satisfies the \citet{Robbins1951} conditions.
\end{itemize}
\begin{figure}[h]
\centering
\includegraphics[scale = 0.4]{gradientascent}
\caption{Red steps show the path a variable $\lambda^{(m)}$ took while finding the maximum of a standard normal distribution.}
\label{fig:GradAscent}
\end{figure}
\end{frame}

\begin{frame}
\section{Copulas}
\frametitle{Vine Copulas}
\begin{itemize}
\item Vine copulas allow distributions to be constructed where a flexible dependence structure is fit independently to the margins.
\vspace{3mm}
\item Problem: There are $O(k!)$ ways to construct a vine copula for a $k-$dimensional parameter vector $\boldsymbol{\theta}$.
\item Solution: Di{\ss}mann's algorithm generates a vine copula that fits a sample of data reasonably well.
\item Implementation: Run MCMC and use Di{\ss}mann's algorithm to find a good vine copula to use as $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$. Run SVB to update this approximation as more data comes in.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Exponential Smoothing}
\begin{itemize}
\item \citet{Taylor2008} provides evidence that a double-seasonal Holt-Winters exponential smoothing model performs better than common alternatives for very short term electricity load forecasting.
\item \citet{Forbes2000} provides a Bayesian approach to estimating exponential smoothing models by Monte Carlo simulation.
\item We slightly modify the Holt-Winters model to work within the Bayesian approach.
\item Applying SVB will allow electricity load density forecasts to be made at a very short time-frame.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Double Seasonal Holt-Winters Exponential Smoothing}
\begin{itemize}
\item We use the modified double seasonal Holt-Winters exponential smoothing model
\begin{align}
y_t &= l_{t-1} - \sum_{i = 1}^{m_1 - 1}d_{t-i} - \sum_{i = 1}^{m_2 - 1}w_{t-i} + e_t \label{ds-hw-rp1} \\
l_t &= l_{t-1} + \alpha e_t \label{ds-hw-rp2} \\
d_t &= - \sum_{i = 1}^{m_1 - 1}d_{t-i} + \delta e_t \label{ds-hw-rp3} \\
w_t &= - \sum_{i = 1}^{m_2 - 1}w_{t-i} + \omega e_t \label{ds-hw-rp4}.
\end{align}
where $m_1$ is the length of the daily cycle, and $m_2$ is the length of the weekly cycle.
\item (\ref{ds-hw-rp3}) and (\ref{ds-hw-rp4}) restrict the seasonality but allow us to avoid inverting a singular matrix in the Bayesian approach.
\end{itemize}
\end{frame}

@@

\begin{frame}
\bibliographystyle{asa}
\bibliography{references}
\end{frame}

>>>>

\newcounter{saveenumi}
\newcommand{\seti}{\setcounter{saveenumi}{\value{enumi}}}
\newcommand{\conti}{\setcounter{enumi}{\value{saveenumi}}}
\resetcounteronoverlays{saveenumi}

@@

\begin{frame}
\tableofcontents
\end{frame}

@@

\begin{frame}
\frametitle{Bayesian Computation Techniques}
\begin{itemize}
\item A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
\begin{enumerate}
\item Converges to the posterior density.
\item Can be very slow.
\item Scales badly to large datasets.
\end{enumerate}
\item Interest is rapidly developing in approximations, such as Variational Bayes (VB). VB:
\begin{enumerate}
\item Is significantly faster than MCMC.
\item Scales to large datasets.
\item Has no theoretical guarantee to converge to the posterior.
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\section{Markov Chain Monte Carlo}
\frametitle{Markov Chain Monte Carlo}
\begin{itemize}
\item A Gibbs MCMC sampler iteratively samples the full conditional distributions:
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
\item MCMC often requires a large number of samples.
\item The predictive density may be unavailable before more data is observed.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Gibbs Full Conditional Distributions}
\begin{itemize}
\item Let $y_t \overset{iid}{\sim} \mathcal{N}\left(\mu, \sigma^2\right) \mbox{ for } t = 1, 2, \dots, T$. 
\item Use priors
\begin{enumerate}
\item $\mu | \sigma \sim \mathcal{N}\left(\gamma, \sigma^2/\tau\right)$
\item $\sigma^2 \sim IG\left(\mbox{shape} = \alpha, \mbox{scale} = \beta\right)$.
\end{enumerate}
\vspace{2mm}
\item Gibbs MCMC samples the full conditional distributions,
\begin{enumerate}
\item $\mu | \sigma^2, y_{1:T} \sim \mathcal{N}\left(\bar{\gamma}, \sigma^2 / \bar{\tau}\right)$
\item $\sigma^2 | \mu, y_{1:T} \sim IG \left(\mbox{shape} = \bar{\alpha}, \mbox{scale} = \bar{\beta} \right)$
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\section{Variational Bayes}
\frametitle{Variational Bayes}
\begin{itemize}
\item Key Idea: Replace the intractable $p(\boldsymbol{\theta} | y_{1:T})$ with a tractable approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
\item How: Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions is closest to $p(\boldsymbol{\theta} | y_{1:T})$ by some metric. 
\item Typically minimise the Kullback-Leibler divergence from $q$ to $p$,

\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \mathbb{E}_q  \left[ \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) \right]
\end{equation}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Evidence Lower Bound}
\begin{itemize}
\item Rewrite (\ref{KL-def}) as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} |y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}),
\end{equation}
where
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \mathbb{E}_{q} \left[ \ln (p(y_{1:T},\boldsymbol{\theta})) - \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) \right].
\end{equation}
\item $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is called the Evidence Lower Bound (ELBO).
\item Maximising the ELBO $\iff$ Minimising the KL Divergence.
\item There are two tasks: 
  \begin{enumerate}
  \item Choose a functional form for $q$. 
  \item Optimise auxillary parameter vector $\boldsymbol{\lambda}$. 
  \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Mean Field Variational Bayes - MFVB}
\begin{itemize}
\item Early Variational Bayes implementation pioneered by Jordan \textit{et al.} (1999) and Ghahramani and Beal (2000).
\item Restricts $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the factorisable family,
\begin{equation}
\label{mf1}
q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
\end{equation}
\item Maximise the ELBO by setting each marginal to
\begin{equation}
\label{mf2}
q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{j \neq i}}[\ln(p(y_{1:T},\boldsymbol{\theta}))]).
\end{equation}
\item Solution only in closed form for exponential family models with limited prior choice.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{$q$ distributions for the Normal-Inverse Gamma model}
\begin{itemize}
\item (\ref{mf2}) is applied to the Normal-Inverse Gamma model to find $q(\mu | \boldsymbol{\lambda}_1)$.
\end{itemize}
\begin{enumerate}
\item Write 
\begin{align}
\log(p(y_{1:T}, \mu, \sigma^2)) &= -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} \nonumber \\
&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2} + c. \nonumber 
\end{align}
\item Ignore components without $\mu$
\begin{equation}
\log(p(\mu | y_{1:T}, \sigma^2)) = - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c. \nonumber
\end{equation}
\seti
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{$q$ distributions for the Normal-Inverse Gamma model}
\begin{enumerate}
\conti
\item Take the expectation with respect to $q(\sigma^2 | \boldsymbol{\lambda}_2)$
\begin{equation}
\log(q(\mu | \boldsymbol{\lambda}_1)) = - 1/2 \mathbb{E}_{q(\sigma^2)} [\sigma^{-2}] \left( \sum_{t}(y_t - \mu)^2 + \tau(\mu - \gamma)^2 \right)  + c. \nonumber
\end{equation}
\item Manipulate until a distribution kernel is recognised. 
\item If unrecognised, add a secondary approximation.
\item Repeat for $q(\sigma^2 | \boldsymbol{\lambda}_2)$.
\end{enumerate}
\begin{itemize}
\item In this case $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim IG(\tilde{\alpha}, \tilde{\beta})$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{$\boldsymbol{\lambda}$ for the Normal-Inverse Gamma model}
\begin{itemize}
\item The ELBO maximising choice of $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$ is:
\begin{align}
\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}[\sigma^{-2}]\right )^{-1} \label{mf6} \\
\tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
\tilde{\beta} &= \beta + 1/2\bigg((\tau + T)\mathbb{E}_{q(\mu)}[\mu^2 ]  \nonumber \\ 
&- 2 \mathbb{E}_{q(\mu)}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \bigg). \label{mf8}
\end{align}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{$\boldsymbol{\lambda}$ for the Normal-Inverse Gamma model}
\begin{itemize}
\item Substitute in expectations to (\ref{mf6}) and (\ref{mf8}):
\begin{align}
\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau \right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
\end{align}
\item Circular dependence in (\ref{mf9}) and (\ref{mf10}) 
\item Requires a coordinate ascent algorithm.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Fit to simulated data}
\begin{itemize}
\item Marginal posterior densities for the Normal Inverse Gamma model with $T = 100, \mu = 2$ and $\sigma^2 = 1$.
\item Hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. 
\end{itemize}
\begin{figure}[h]
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{norminvg}
\caption{The analytical true density (black), the Monte Carlo (red) and the MFVB approximation (blue). MFVB converged in same time as required to take three MCMC samples of each parameter.}
\label{fig:norminvg}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Stochastic Variational Bayes - SVB}
\begin{itemize}
\item Paisley \textit{et al.} (2012) and Ranganath \textit{et al.} (2014) adapted a gradient ascent algorithm to maximise the ELBO.
\item Gradient ascent returns the optimal $\boldsymbol{\lambda}$ for a specified functional form $q$.
\item Only requires Monte-Carlo estimates of $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$,
\begin{align}
\label{SVB}
\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) &\approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda})] \nonumber \\
&\times \big(\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}))\big) 
\end{align}
where $\boldsymbol{\theta}_s$ is simulated from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ for $s = 1, 2, \dots, S$. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Gradient Ascent}
\begin{itemize}
\item Make steps of the form
\begin{equation}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} f(\lambda) \bigg|_{\lambda = \lambda^{(m)}}
\end{equation}
until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$, with $\rho^{(m)}, m = 1, 2, \dots$ that satisfies the Robbins and Munro (1951) conditions.
\end{itemize}
\begin{figure}[h]
\centering
\includegraphics[scale = 0.4]{gradientascent}
\caption{Red steps show the path a variable $\lambda^{(m)}$ took while finding the maximum of a standard normal distribution.}
\label{fig:GradAscent}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Choosing the $q$ distribution}
\begin{itemize}
\item Finding a good functional form for $q$ is an open problem.
\item Approaches include heirarchical distributions and neural network transformations.
\item $q$ on needs to be parsimonous to scale.
\item We augment $q$ with a copula to model dependence separately to the margins.
\end{itemize}
\end{frame}

\begin{frame}
\section{Copulas}
\frametitle{Vine Copulas}
\begin{itemize}
\item Vine copulas increase the flexiblity of copula modelling.
\vspace{3mm}
\item Problem: There are $O(k!)$ ways to construct a vine copula for a $k-$dimensional parameter vector $\boldsymbol{\theta}$.
\item Solution: Di{\ss}mann's algorithm generates a vine copula that fits a sample of data reasonably well.
\item Implementation: Run MCMC and use Di{\ss}mann's algorithm to find a good vine copula to use as $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$. Run SVB to update this approximation as more data comes in.
\end{itemize}
\end{frame}

@@

@@
